Estimating mapping quality when using seeding heuristics for DNA alignment

### 2019-03-31 MEM.tex
First complete sketch with exact, skip and MEM seeds.

\title{Estimating mismapping rates in seed-based algorithms
for read mapping}

@@

We will refer to this question as the \emph{true location problem}.
It has applications in countless domains, like discovering
disease-causing mutations; diagnosing viral, bacterial or fungal
infections; diagnosing cancer and choosing appropriate therapies;
selecting breeds of interest in agriculture; tracing human migrations from
ancient bones; identifying the donor of a biological sample; diagnosing
genetic diseaes at birth; finding compatible graft donors \textit{etc}.

@@

Because sequencing errors can convert a DNA fragment into one of its
duplicates, the \emph{true} location of a DNA fragment is not always the
\emph{best} (as measured by the identity between the sequence and the
candidate location). This leads to the following question: how can we
identify the \emph{best} location of the fragment in the genome?

@@

Importantly, the trade-off depends on the seeding method. This means that
we can develop faster mapping algorithms at no cost for accuracy, as long
as we can find better seeding strategies. Progress on this line of
research has largely benefited from the improvement of computer hardware
and from the deveopment of indexing algorithms. Literature on this topic
is already abudant. Some examples of seeding algorithms are given in
references~\cite{sun2005designing,pmid11934743,xu2006optimizing,
pmid17044164,brejova2003vector,pmid18684737,pmid15359419}.
References~\cite{pmid16533404} and~\cite{pmid20460430} present high-level
comparisons of different designs, and reference~\cite{navarro2001guided}
gives a global overview of filtration methods in pattern matching.

\subsection{The two types of seeding failure}

@@

\begin{definition}
The seeding process is said to be
\begin{enumerate}[i)]
\item on target if the candidate set contains the target,
\item off target if the candidate set contains a duplicate but
not the target, 
\item null if the candidate set contains neither.
\end{enumerate}
\end{definition}

If a seed-based algorithm returns an incorrect location, then the best
location was never tested with an exact alignment. If the error is given a
high confidence, then it must be the best in the candidate set and it must
be similar to the read. This narrows down the cases of mismapping, or more
exactly high confidence mismapping to off-target seeding.

This observation is fundamental. In order to estimate the mismapping rate
of an algorithm, we must know how often the associated seeding heuristic
is off target. Yet, this information is missing for nearly all seeding
methods, meaning that current mappers cannot be properly calibrated.

The focus of this article is to develop a theory to estimate the
probability of off-target seeding under the most popular seeding
heuristics. Our previous work pioneered a method to compute seeding
probabilities but it did not distinguish off-target from null
seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
not provide a way to control the mismapping rate. Other authors
investigated the reliability of mapping algorithms~\cite{pmid23872968},
but they focused on the probability that random sequences have a single
hit, recognizing that solving the problem requires taking into
consideration the repeat structure of the genome.

\section{Types of seeds}

The term ``seed'' does not have exactly the same meaning according to
different authors. It can designate a part of the read, a part of the
genome, a particular sequence motif or a structured pattern of matches.
Also, the term is not always short for ``exact match''. For instance, the
algorithm PatternHunter~\cite{pmid11934743} describes ``spaced seeds''
that tolerate mismatches\footnote{For the anecdote, the method is covered
by a patent (reference US20070088510A1).}. To avoid potential confusion,
we will adopt the following convention.

@@

When a seed matches a particular location of the genome, we will say that
it is a seed \emph{for} that location. For instance, we will often refer
to a seed for the target or a seed for one of its duplicates.

This definition presents a computational challenge: to know if a given
subsequence is a seed we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to solve it, even when the reference genome is
very large~\cite{ferragina2000opportunistic, ferragina2005indexing}. Such
algorithms are crucial for the present theory, but describing them
falls outside the scope of the document. Looking ahead to
section~\ref{sec:est}, let us only mention that this can be done with an
algorithm called the \emph{backward
search}~\cite{ferragina2000opportunistic}. We refer the interested reader
to the relevant
literature~\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736} and
here on we will assume that seeds are always known without ambiguity [ALL
THE CITED LITERATURE IS FOR MEM SEEDS, WE NEED TO FIND ARTICLES THAT TREAT
THE CASE OF OTHER SEEDS].

@@

Exact seeds have a fixed size $\gamma$. When using exact seeds, all the
perfect matches of size $\gamma$ are collected to constitute the candidate
set. This seeding heuristic was used in the first version of
BLASTN~\cite{pmid2231712}, but is no longer used because they produce many
overlapping matches in the same regions, as we are about to illustrate.

Fig.~\ref{fig:exact_seed_example} shows the example of a read produced
by an imperfect sequencer. The sequenced DNA fragment comes from a region
of the genome that has three extra duplicates, and the read has three
miscalled nucleotides (indicated by stars).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{exact_seed_example.pdf}
\caption{\textbf{Example of exact seeds.}
A sequenced DNA fragment (Read) is shown above its original form in the
genome (Target).}
\label{fig:exact_seed_example}
\end{figure}

Fig.~\ref{fig:exact_seed_example} also illustrates the seeding process
with exact seeds of size $\gamma=7$. It is clear that erroneous
nucleotides can also belong to exact seeds because they sometimes match
one of the duplicates. However, this is not the case of erroneous
nucleotides that are also mismatches for \emph{all} the duplicates, like
the second sequencing error in this example.

Thie middle of the read shows some cluttering where consecutive seeds
match consecutive sequences at the same location. This is typical for
exact seeds and constitutes a nuisance for the implementation. Indeed, it
is a waste of computer resources to discover matches in sequences that are
already in the candidate set. In addition, this seeding method is not
particularly sensitive~\cite{pmid11934743} so it is used only for specific
applications. Nevertheless, it will be extremely useful for the
development of the present theory.

@@

Skip seeds have a fixed size $\gamma$, but unlike exact seeds they cannot
start at every nucleotide. Instead, a fixed amount of nucleotides is
skipped between every seed. This is a way to reduce the amount of
overlapping matches at the same location, but the cost is a loss of
sensitivity. This seeding heuristic is used by
Bowtie2~\cite{pmid22388286}, where seeds have size $\gamma=16$ and fall
every 10 nucleotides (9 nucleotides are skipped between seeds).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{skip_seed_example.pdf}
\caption{\textbf{Example of skip seeds.}
The sequences and the annotations are the same as in
Fig.~\ref{fig:exact_seed_example}, but here one nucleotide is skipped
betwen the seeds. In other words, seeds can never start on a nucleotide
with an even index. To highlight the difference with
Fig.~\ref{fig:exact_seed_example}, the missing seeds are represented by
dotted lines. Skipping nucleotides reduces the clutter at the center of
the read, but it also removes the only seed for the second duplicate.}
\label{fig:skip_seed_example}
\end{figure}

Fig.~\ref{fig:skip_seed_example} shows the seeding process on the read of
Fig.~\ref{fig:exact_seed_example} when using skip seeds instead of exact
seeds. Here the size is still $\gamma=7$ but 1 extra nucleotide is skipped
between seeds. This amounts to removing every other seed. The consequence
is that there are fewer overlapping matches at the center of the read, but
the only seed for the second duplicate was lost. This is a rather positive
outcome because there is one off-target location fewer in the candidate
set, but in general the same might happen to the target.

It is clear that skipping nucleotides reduces the sensitivity of the
seeding, but to what extent? One could test this empirically, but it
depends on the seed length, the amount of nucleotides that are skipped,
the error rate of the sequencer and the size of the reads. The present
theory will allow us to get the answer automatically.

@@

We will now focus on a seeding heuristic known as ``MEM seeding'' (where
MEM stands for Maximal Exact Match). This approach gives good empirical
results when mapping short reads (it is for instance the
strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
Exact Match seeds).} of the popular mapper BWA-MEM~\cite{li2013aligning})
but there is presently no theory to justify its performance. A deeper
understanding of MEM seeding would be useful to calibrate the
speed/accuracy trade-off of this heurisitic.

Before developing a formal theory, let us explain in more detail what
Maximal Exact Matches are and how MEM seeding works.

\begin{definition}
A Maximal Exact Match (MEM) seed is a subsequence of a read that is
present in the genome of interest and that cannot be extended --- either
because the read ends or because the extended subsequence is not in the
genome.
\end{definition}

Fig.~\ref{fig:MEM_example} shows a concrete example where a read is
aligned to candidate locations. The target (the true location of the
read) is first potential location. The locations below correspond to
duplicate sequences. The mismatches between the genome and the read are
highlighted in bold. The read has three errors indicated by stars, so
\texttt{T}, \texttt{T} and \texttt{C} should actually be \texttt{A},
\texttt{A} and \texttt{G} (\textit{i.e.} as in the target sequence).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{MEM_example.pdf}
\caption{\textbf{Example of MEM seeds.}
A DNA fragment is sequenced with an imperfect instrument. It comes from
the top genomic location (Truth), but three errors were introduced
(indicated by stars). The sequence of the DNA fragment has three
duplicates at other genomic locations, shown below the target. Mismatches
between the read and the candidate sequence are highlighted in bold. MEM
seeds are indicated by grey boxes.}
\label{fig:MEM_example}
\end{figure}

The MEM seeds are indicated by grey boxes. Technically, they are part of
the read, but they are shown on the genome to highlight the position of
the match. One MEM seed cannot cover a smaller one (otherwise the smaller
would not be a MEM seed), but two consecutive MEM seeds can overlap, as in
the case of the frst two from the left. Note that overlapping MEM seeds
must match distinct sequences of the genome (otherwise neither of them
would be a MEM seed). Consecutive MEM seeds can also be disjoint, as in
the case of the second and the third from the left. In this case, they
must be separated by a nucleotide that is a mismatch against \emph{all}
the duplicates.

\section{Model and computational strategy}
\label{sec:symbolic}

\subsection{Sequencing error and duplications}

Now that this is clarified, we need to model the sequencing and the
duplication processes so that we can compute the probabilities of the
events of interest. We assume that the sequencing instrument has a
constant substitution rate $p$, and that insertions and deletions never
occur. When a substitution occurs, we assume that the instrument is
equally likely to output any one of the remaining three nucleotides. This
is a coarse-grain description of the error spectrum of the Illumina
technology~\cite{pmid21576222}.

Next, we assume that the target sequence has $N \geq 0$ duplicates, so
that there are $N$ off-target sequences. We further assume that
duplication happened instantaenously at some point in the past and that
all $N+1$ sequences diverge independently of each other at a constant
rate. In other words, we ignore the complications due to the genealogy of
the duplication events and we simply assume that at each position, any
given duplicate is identical to the target with probability $1-\mu$. If it
is not, we assume that the duplicate has any of the remaining three
nucleotides with equal probability (\textit{i.e.} each is found with
probability $\mu/3$).

Note that read errors are always mismatches against the target (because we
assume that the target is the true sequence), and they match each
off-target sequence with probability $\mu/3$. Correct nucleotides are
always a match for the target, and they match each off-target sequence
with probability $1-\mu$.

@@

The central object of analytic combinatorics is the generating function,
and for our particular purpose we will use a special kind known as
\emph{weighted generating function}.

\begin{definition}
\label{def:GF}
Let $\mathcal{A}$ be a set of combinatorial objects characterized by a
size in $\mathbb{N}$ and a weight in $\mathbb{R}^+$. The weighted
generating function of $\mathcal{A}$ is defined as

\begin{equation}
\label{eq:GF1}
A(z) = \sum_{a \in \mathcal{A}} w(a) z^{|a|},
\end{equation}
where $|a|$ and $w(a)$ denote the size and weight of the object $a$,
respectively. Expression (\ref{eq:GF1}) also defines a sequence $(a_k)_{k
\geq 0}$ such that

@@

To give a concrete example, if the probability of the symbol
$\downarrow_{/0}$ is $q$, the weighted generating function of a segment
containing only this symbol is $qz$. In our case, we are interested in the
weighted generating function of reads without good seeds, so that the
coefficient $a_k$ is equal to the probability that a read of size $k$ has
no good seed.

@@

The symbolic method is a strategy belonging to the general \emph{analytic
combinatorics} framework. The idea is to combine simple combinatorial
objects into more complex objects. Each combinatorial operation on the
objects corresponds to a mathematical operation on their weighted
generating functions. One can thus obtain the weighted generating function
of complex objects, and from there extract their probability of
occurrence.

The combinatorial construction of reads using the extended MEM alphabet
gives us a method to express their weighted generating function from the
simpler generating function of segments. More specifically, we can
describe complex sequences of segments through a so-called \emph{transfer
matrix} $M(z)$ that specifies which pairs of consecutive segments are
allowed and which are forbidden in the reads of interest.

The segment types are identified by their terminators arranged in a
predefined oder. The entry of $M(z)$ at coordinates $(i,j)$ is the
weighted generating function of segments with the $j$-th terminator that
can be inserted immediately after segments with the $i$-th terminator. The
entries of the matrix $M(z) + M(z)^2 + \ldots = M(z) \cdot (I-M(z))^{-1}$
are the weighted generating functions of the reads of all sizes using all
the allowed combinations of segments.

The entry of interest from $M(z)$ is the one that corresponds to the
terminators $\downarrow_{/0}$ and $|$ because it represents the reads of
any size whose first segment can be inserted after $\downarrow_{/0}$, and
whose last segment is terminated by $|$. Indeed, the $|$ terminator marks
the end of the read, so it must be the final symbol of every valid read.
As for the first segment, observe that all the segments that can follow
the $\downarrow_{/0}$ symbol are exactly those that can be at the frst
postion of the read\footnote{One could have introduced a special start
symbol. It is equivalent to an invisible $\downarrow_{/0}$ prepending
the reads.}.

In summary, the transfer matrix $M(z)$ contains the weighted generating
function of interest. Our next task is thus to specify $M(z)$.

More didactic descriptions of the strategy to compute seeding
probabilities using analytic combinatorics have been published
elsewhere~\cite{filion2017analytic,filion2018analytic}. The interested
readers can also discover more about analytic combinatorics in the
standard textbooks~\cite{flajolet2009analytic,sedgewick2013introduction}.

\subsection{Example 1: on-target exact seeds}
\label{sec:sample_exact}

We highlight the strategy above with an example that will turn out to be
central for the development of the theory. In addition, it is simple
enough to provide a gentle introduction to the general methodology. This
example was described in detail in~\cite{filion2017analytic} and
in~\cite{filion2018analytic}, but we repeat it here with a formalism that
fits the present manuscript.

The first step is to note that the nucleotide sequence of the read is
irrelevant. Indeed, the read has an on-target seed if and only if it
contains a stretch of $\gamma$ nucleotides without error. So the main step
is to recode the read as sequence of correct or erroneous nucleotides.

We define the mismatch alphabet $\mathcal{A}_0 = \{\square, |,
\Downarrow,\}$, where $\square$ represents a correct nucleotide,
$\Downarrow$ represents an erroneous nucleotide and $|$ is special symbol
appended after the last nucleotide to mark the end of the read.

This recoding allows us to segment the read in \emph{segments} defined as
a sequence of 0 or more symbols $\square$ followed by a symbol
$\Downarrow$. The last part of the read is called the tail and it is
terminated by the symbol $|$ instead of $\Downarrow$. The symbol $|$ does
not correspond to a nucleotide, it has size 0. This decomposition is
unique, so we can view a read as a sequence of segments with a tail,
instead of a sequence of nucleotides.

Fig.~\ref{fig:simple} shows an example of decomposition in segments.
On-target seeds cannot contain sequencing errors, therefore they must lie
inside a segment or inside the tail. So the size of the segments and of
the tail indicates whether the read contains an on-target seed.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_simple.pdf}
\caption{\textbf{The mismatch encoding.}
A example read is represented in the match alphabet. The symbol
$\Downarrow$ represents a mismatch against the target sequence and
the symbol $\square$ represents a match.}
\label{fig:simple}
\end{figure}

>>>>

\title{Estimating mapping quality when using seeding heuristics for
DNA alignment}

@@

We will refer to this question as the \emph{true location problem}.
It has applications in countless domains, like discovering
disease-causing mutations; diagnosing viral, bacterial or fungal
infections; diagnosing cancer and choosing appropriate therapies;
selecting breeds of interest in agriculture; tracing human migrations from
ancient bones; identifying the donor of a biological sample; diagnosing
genetic diseases at birth; finding compatible graft donors \textit{etc}.

@@

Since repeats play a central role in the problem at hand, we give the
terms ``targets'' and ``duplicates'' a meaning that will facilitate the
exposition of the theory.

\begin{definition}
The target is the DNA fragment that was sequenced. Duplicates are
sequences of the genome that are homologous to the target.
\end{definition}

Because sequencing errors can convert the sequence of the target into one
of its duplicates, the \emph{true} location of a DNA fragment is not
always the \emph{best} (as measured by the identity between the sequence
and the candidate location). This leads to the following question: how can
we identify the \emph{best} location of the fragment in the genome?

@@

Importantly, the trade-off depends on the seeding method. This means that
we can develop faster mapping algorithms at no cost for accuracy, as long
as we can find better seeding strategies. Progress on this line of
research has largely benefited from the improvement of computer hardware
and from the development of indexing algorithms. Literature on this topic
is already abundant. Some examples of seeding algorithms are given in
references~\cite{sun2005designing, pmid11934743, xu2006optimizing,
kucherov2005multiseed, brejova2003vector, pmid18684737, pmid15359419}.
References~\cite{pmid16533404} and~\cite{pmid20460430} present high-level
comparisons of different designs, and reference~\cite{navarro2001guided}
gives a global overview of filtration methods in pattern matching.

\subsection{The two types of seeding failure}
\label{sec:twotypes}

@@

\begin{definition}
The seeding process is said to be
\begin{enumerate}[i)]
\item ``on target'' if the candidate set contains the target,
\item ``off target'' if the candidate set contains a duplicate but
not the target, 
\item ``null'' if the candidate set contains neither.
\end{enumerate}
\end{definition}

If a seed-based mapping algorithm returns an incorrect location, then it
must be the case that the best location was never tested in the second
phase, otherwise it would have been singled out by the exact alignment
algorithm.

This observation is fundamental: in order to estimate the mismapping rate
of an algorithm, we must know how often the associated seeding heuristic
is off target. For lack of theoretical framework, this information is
missing for nearly all existing seeding methods, meaning that current
mappers cannot be properly calibrated.

The focus of this article is to develop a theory to estimate the
probability that seeding is off target under the most popular seeding
heuristics for short reads. Our previous work pioneered a method to
compute seeding probabilities but it did not distinguish off-target from
null seeding~\cite{filion2017analytic,filion2018analytic}, and therefore
did not provide a way to control the mismapping rate. Other authors
investigated the reliability of mapping algorithms~\cite{pmid23872968},
but they focused on the probability that random sequences have a single
hit, recognizing that solving the problem requires taking into
consideration the repeat structure of the genome. Here we address this
problem in general terms and we provide a practical way to estimate the
mismapping rate in complex genomes.

The rest of this document is organized to present the content in a
didactic manner. The order of the examples is meant to facilitate the
aquisition of the main concepts, meaning that we sometimes go back and
forth between different types of seeding problems. We first give some
background on the different types of seeding methods considered here, we
then introduce the computational strategy based on transfer matrices, and
then delve into each specific seeding method. Finally, we provide a
statistical method to estimate the parameters of the problem from the
data.

\section{Seeds}

The term ``seed'' does not have exactly the same meaning according to
different authors. It can designate a part of the read, a part of the
genome, a particular sequence motif or a structured pattern of matches.
Also, the term does not always imply an exact match. For instance, the
algorithm PatternHunter~\cite{pmid11934743} uses ``spaced seeds''
that tolerate mismatches\footnote{Spaced seeds are covered
by patent US20070088510A1.}. To avoid potential confusion,
we will adopt the following convention.

@@

When a seed matches a particular location of the genome, we will say that
it is a ``seed for that location''. For instance, we will often use
expressions such as ``a seed for the target'' or ``a seed for a
duplicate''.

This definition presents a computational challenge: to know if a given
subsequence is a seed we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to solve it, even when the reference genome is
very large.

These algorithms are crucial for the present theory, but describing them
in depth is outside the scope of this document. Let us just mention that
all the methods belong to a family known as exact offline string matching
algorithms, where ``offline'' means that sequences are looked up in an
index instead of the genome itself. Online methods can be used when the
reference genome is not indexed~\cite{faro2013exact}, but this case is of
little relevance in the present context.

The index is usually a hash table or a variant of the so-called
FM-index~\cite{pmid20460430}. Hash tables are typically used to index
$k$-mers, which makes them useful to search seeds of set size $k$,
(see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
algorithms). In contrast, the FM-index has no set size so it can be used
to search seeds of different lengths~\cite{ferragina2000opportunistic,
ferragina2005indexing}. It is a compact data structure based on the suffix
array~\cite{manber1993suffix} and on the Burrows-Wheeler
transform~\cite{burrows1994block}, emulating a suffix trie with a much
smaller memory footprint. Queries in the FM-index are performed with an
algorithm known as the \emph{backward
search}~\cite{ferragina2000opportunistic}.

Other methods can be efficient (\textit{e.g.}, running a bisection on the
suffix array~\cite{dobin2013star}) but the FM-index is currently the most
popular choice for seeding methods. For MEM seeds described below, it is
even the only practical option~\cite{pmid24336412, pmid25399029,
pmid23349213, pmid19389736}, so we implicitly assume that seeding is
performed by running the backward search. This will only have practical
consequences in section~\ref{sec:est}, where we propose a method to
estimate the parameters that are required to compute the probabilities of
interest. For the rest, we simply assume that seeds are known at all times
without ambiguity because this problem has several practical solutions.

@@

Exact seeds have a fixed size $\gamma$. When using exact seeds, all the
perfect matches of size $\gamma$ are collected to constitute the candidate
set. This seeding heuristic was used in the first version of
BLASTN~\cite{pmid2231712}, but is no longer used because it produces many
overlapping matches in the same regions, as we are about to illustrate.

Fig.~\ref{fig:exact_seed_example} shows the example of a read produced
by an imperfect sequencer. The sequenced DNA fragment comes from a region
of the genome that has three duplicates (there are four homologous
sequences in total), and the read has three miscalled nucleotides
(indicated by stars).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{exact_seed_example.pdf}
\caption{\textbf{Example of exact seeds.}
A sequenced DNA fragment (Read) is shown above the actual molecule
(Target), and above homologous sequences also present in the genome
(Duplicates). Sequencing errors are indicated by a star, each marked
\texttt{T} was actually an \texttt{A} that was misread. The mismatches
between the genomic sequences and the read are indicated in bold. The
exact seeds of size $\gamma = 7$ are indicated as horizontal grey lines
above the read. The matching regions in the genomic sequences are
shadowed in grey. Several overlapping seeds accumulate at the center of
the read, which is typical for this seeding method.}
\label{fig:exact_seed_example}
\end{figure}

The exact seeds with $\gamma=7$ are shown above the sequence of the read.
Observe that erroneous nucleotides can belong to exact seeds because they
sometimes match one of the duplicates. For instance, the first sequencing
error matches all the duplicates and belongs to a seed for the first one.
However, erroneous nucleotides that are mismatches for \emph{all} the
duplicates cannot belong to a seed. This is the case of the second
sequencing error in this example, where there is a local deficit of
seeds.

Thie middle of the read shows some clutter where consecutive seeds
match consecutive sequences at the same location. This is typical for
exact seeds and constitutes a nuisance for the implementation. Indeed, it
is a waste of computer resources to discover matches in sequences that are
already in the candidate set. In addition, this seeding method is not
particularly sensitive~\cite{pmid11934743} so it is used only for specific
applications. Nevertheless, it will be extremely useful for the
development of the present theory.

@@

Skip seeds have a fixed size $\gamma$, but unlike exact seeds they cannot
start at every nucleotide. Instead, a fixed amount of nucleotides is
skipped between every seed. This is a way to reduce the amount of
overlapping matches at the same location, at the cost of sensitivity. This
seeding heuristic is the core of Bowtie2~\cite{pmid22388286}, where seeds
have size $\gamma=16$ and are separated by 10 nucleotides (9 positions are
skipped). We will refer to seeds where $n$ nucleotides are skipped as
``skip-$n$ seeds''. For instance, Bowtie2 uses skip-9 seeds.

\begin{figure}[h]
\centering
\includegraphics[scale=1]{skip_seed_example.pdf}
\caption{\textbf{Example of skip seeds.}
The sequences and the annotations are the same as in
Fig.~\ref{fig:exact_seed_example}, but here we use skip-1 seeds. In other
words, seeds can never start at nucleotides 2, 4, 6 \textit{etc}. To
highlight the difference with Fig.~\ref{fig:exact_seed_example}, the
missing seeds are represented by dotted lines. Skipping nucleotides
reduces the clutter at the center of the read, but it also removes the
only seed for the second duplicate.}
\label{fig:skip_seed_example}
\end{figure}

Fig.~\ref{fig:skip_seed_example} shows what happens when exact seeds are
replaced by skip-2 seeds on the read of Fig.~\ref{fig:exact_seed_example}.
Here the size is still $\gamma=7$ but 1 nucleotide is skipped between
seeds. This amounts to removing every second seed. The consequence is that
there are fewer overlapping matches at the center of the read, but the
only seed for the second duplicate is lost. This is a rather positive
outcome because there is one off-target location fewer in the candidate
set, but the same might happen to the target.

It is clear that skipping nucleotides reduces the sensitivity of the
seeding step, but to what extent? One could test this empirically, but the
answer depends on the seed length, the number of nucleotides that are
skipped, the error rate of the sequencer and the size of the reads. The
present theory will allow us to make general statements about the
performance of skip seeds against exact seeds in different contexts.

@@

MEM seeds (where MEM stands for Maximal Exact Match) are somewhat harder
to define. Unlike exact seeds and skip seeds, their size is variable as we
are about to see. They are used in BWA-MEM~\cite{li2013aligning} where
they give good empirical results (but note that the actually seeding
scheme uses more complex Super Maximal Exact Matches). To describe MEM
seeds, let us first focus on MEMs.

\begin{definition}
A Maximal Exact Match (MEM) is a subsequence of the read that is present
in the reference genome and that cannot be extended --- either because the
read ends or because the extended subsequence is not in the genome.
\end{definition}

A \emph{MEM seed} is simply a MEM of minimum size $\gamma$.
Fig.~\ref{fig:MEM_example} shows what happens when exact seeds or skip
seeds are replaced by MEM seeds on the read of
Fig.~\ref{fig:exact_seed_example}. Observe that the clutter at the center
of the read has disappeared because consecutive matches are fused into a
single MEM seed.

\begin{figure}[h]
\centering
\includegraphics[scale=1]{MEM_example.pdf}
\caption{\textbf{Example of MEM seeds.}
The sequences and the annotations are the same as in
Fig.~\ref{fig:exact_seed_example}, but here we use MEM seeds of minimum
size $\gamma=7$. The clutter at the center of the read has disappeared and
there is at least one seed for each sequence.}
\label{fig:MEM_example}
\end{figure}

Two consecutive MEM seeds can overlap, in which case they always match
distinct sequences of the genome (otherwise neither of them would be a MEM
seed). There does not have to be any overlap though, because a nucleotide
can be a mismatch against \emph{all} the sequences, like the second read
error for instance.

Note that a MEM does not always match a single sequence of the genome. For
instance, the rightmost MEM seed matches two distinct genomic sequences.
This case motivates the following definition, which will play an important
role later.

Compared to seeds of fixed size, MEM seeds have two counter-intuitive (and
undesirable) properties. The first is that there are cases where there
cannot be any good seed, even when changing the sensitivity parameter
$\gamma$. Fig.~\ref{fig:full_masking_example} shows such an example. Even
though there is a single sequencing error, the read has no MEM seed for
the target. Lowering $\gamma$ does not change this, so there is no way to
discover the true location in this case (even though it is the best
location).

The second counter-intuitive property is that shortening a read can
sometimes generate a MEM seed for the target. Fig.~\ref{fig:short_vs_long}
shows an example of this case. There is no MEM seed for the target, but
there would be one if the read were two nucleotides shorter on the right
side. Indeed, in this case there would be a shared MEM seed matching the
first and the second locations (provided $\gamma \leq 12$).

These examples show that MEM seeds can perform worse than seeds of fixed
size. MEM seeds yield fewer candidates and therefore speed up the mapping
process, but the question is at what cost for accuracy? The theory
developed here will allow us to compute the probability that a read has
no MEM seed for the target and thus that the true location is missed at
the seeding stage.

\section{Model and strategy}
\label{sec:symbolic}

\subsection{Sequencing errors and divergence of the duplicates}
\label{sec:error}

We now need to model the sequencing and duplication processes so that
we can compute the probabilities of the events of interest. We assume that
the sequencing instrument has a constant substitution rate $p$, and that
insertions and deletions never occur. When a substitution occurs, we
assume that the instrument is equally likely to output any one of the
remaining three nucleotides. This corresponds more or less to the error
spectrum of the Illumina sequencing technology~\cite{pmid21576222}.

Next, we assume that the target sequence has $N \geq 0$ duplicates, so
that there are $N$ off-target sequences. We further assume that
duplication happened instantaenously at some point in the past and that
all $N+1$ sequences diverge independently of each other at a constant
rate. In other words, we ignore the complications due to the genealogy of
the duplication events and we simply assume that at each nucleotide
position, any given duplicate is identical to the target with probability
$1-\mu$. If it is not, we assume that the duplicate sequence can be any of
one the remaining three nucleotides (\textit{i.e.} each is found with
probability $\mu/3$).

Note that read errors are always mismatches against the target (because we
assume that the target is the true sequence), and they match each
off-target sequence with probability $\mu/3$. In contrast, correct
nucleotides are always matches for the target, and they match each
off-target sequence with probability $1-\mu$.

Before going further, we also need to move a practical consideration out
of the way. Seeds can match any sequence of the genome, not just the
target or the duplicates. However, we will ignore matches in the rest of
the genome because such random matches are unlikely to cause a mapping
error when seeding is off target, contrary to matches in duplicates.
Neglecting those will greatly simplify the exposition of the theory
without loss of generality. We will explain in
section~\ref{sec:short_MEMs} how to deal with this practical case and and
how to identify those matches as off target. Until then, we will consider
that the target and the duplicates are the only sequences in the reference
genome.

@@

The central object of analytic combinatorics is the generating function,
and for our purpose we will use a special kind known as \emph{weighted
generating function}.

\begin{definition}
\label{def:GF}
Let $\mathcal{A}$ be a set of combinatorial objects such that $a \in
\mathcal{A}$ has a size $|a| \in \mathbb{N}$ and a weight $w(a) \in
\mathbb{R}^+$. The weighted generating function of $\mathcal{A}$ is
defined as

\begin{equation}
\label{eq:GF1}
A(z) = \sum_{a \in \mathcal{A}} w(a) z^{|a|},
\end{equation}
Expression (\ref{eq:GF1}) also defines a sequence $(a_k)_{k \geq 0}$ such
that

@@

To give an example, assume that a particular symbol, say $\Downarrow$, has
a probability of occurence equal to $p$. The weighted generating function
of the word containing only this symbol is $pz$. The weight of the word is
its probability (here equal to $p$) and the size is its length in number
of symbols (here 1).

In this document we will focus on the weighted generating function $A(z)$
of the set $\mathcal{A}$ of reads that do not have on-target seeds
(\textit{i.e.} reads for which seeding is either off target or null). The
weight of a read is its probability of occurrence and the size $k$ is its
number of nucleotides. The coefficient $a_k$ is thus the proportion of
reads of size $k$ that do not have an on-target seed, which is the
quantity of interest.

@@

The analytic combinatorics framework relies on a strategy referred to as
the \emph{symbolice method}~\cite{sedgewick2013introduction}. The idea is
to combine simple objects into more complex objects. Each combinatorial
operation on the objects corresponds to a mathematical operation on their
weighted generating functions. One can thus obtain the weighted generating
function of complex objects, whose coefficients $a_k$ are the
probabilities of interest.

As explained in~\cite{filion2017analytic} and~\cite{filion2018analytic},
we recode the reads in alphabets of custom symbols and we specify a
construction plan of the reads using a so-called \emph{transfer matrix}
$M(z)$. The transfer matrix specifies which types of segments can follow
each other in the reads of interest: the entry at coordinates $(i,j)$ is
the weighted generating function of segments of type $j$ that can be
appended to segments of type $i$.

The entry of $M(z) + M(z)^2 + M(z)^3 + \ldots = M(z) \cdot (I-M(z))^{-1}$
at coordinates $(i,j)$ is the weighted generating function of all the
reads that end with a segment of type $j$ and that can be appended to a
segment of type $i$. If segments of types $i$ and $j$ are exactly those
that can be placed at the beginning and at the end of the reads, then this
entry is the weighted generating function of all the reads without
on-target seed. The examples below will clarify the key steps of this
strategy.

A complete description of the strategy to compute seeding probabilities
with the symbolic method can be found in~\cite{filion2017analytic,
filion2018analytic}. The interested readers can also read more about
analytic combinatorics in the popular
textbooks~\cite{flajolet2009analytic, sedgewick2013introduction}.

\subsection{Example 1: on-target exact seeds}
\label{sec:example_exact}

We highlight the strategy above with an example that will turn out to be
central for the development of the theory. In addition, it is simple
enough to provide a gentle introduction to the general methodology. This
example was described in detail in~\cite{filion2017analytic} and
in~\cite{filion2018analytic}, but we repeat it here with a different
formalism to fit the present manuscript.

The first step is to note that the nucleotide sequence of the read is
irrelevant. Indeed, the read has an on-target seed if and only if it
contains a stretch of $\gamma$ nucleotides without error. For this reason,
we recode reads as sequences of correct or erroneous nucleotides.

We define the mismatch alphabet $\mathcal{A}_0 = \{\square, |,
\Downarrow\}$, where $\square$ represents a correct nucleotide,
$\Downarrow$ represents an erroneous nucleotide and $|$ is special symbol
appended after the last nucleotide to mark the end of the read.

This recoding allows us to partition the read in \emph{segments} defined
as a sequence of 0 or more symbols $\square$ followed by a symbol
$\Downarrow$. The last part of the read is called the tail and it is
terminated by the symbol $|$ instead of $\Downarrow$. The symbol $|$ does
not correspond to a nucleotide so it has size 0. Since the decomposition
is unique, we can view a read as a sequence of segments with a tail,
instead of a sequence of nucleotides.

Fig.~\ref{fig:simple} shows an example of decomposition in segments.
On-target seeds cannot contain sequencing errors, therefore they must be
completely embedded inside a segment or inside the tail. So the size of
the segments and of the tail indicates whether the read contains an
on-target seed or not.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_simple.pdf}
\caption{\textbf{The mismatch encoding.}
An example read is represented in the mismatch alphabet. The symbol
$\Downarrow$ represents a mismatch against the target (it is an erroneous
nucleotide) and the symbol $\square$ represents a match (it is a correct
nucleotide). The symbolic sequence of the target is represented below,
where an open square stands for a match and a closed square stands for a
mismatch.}
\label{fig:simple}
\end{figure}

The probability of occurrence of the symbol $\Downarrow$ is $p$, the error
rate of the sequencer, and the probability of occurrence of the symbol
$\square$ is thus $1-p = q$. Both symbols have size 1, so their respective
weighted generating function are $pz$ and $qz$. Using the rule for
concatenation, we see that the weighted generating function of a segment
with $i$ symbols $\square$ is $(qz)^ipz$. The symbol $|$ has size 0, so
the weighted generating function of a tail segment with $i$ symbols
$\square$ is $(qz)^i$.
